---
title: "ST558 Project 1"
author: "Tyler Pollard"
date: "6/15/2021"
output: 
  github_document:
    toc: TRUE
    toc_depth: 3
---

```{r setup, echo = FALSE, eval = TRUE, warning = FALSE, message = FALSE}
library(knitr)
opts_chunk$set(echo = FALSE, eval = TRUE, warning = FALSE, message = FALSE)
#github_document(toc = TRUE, toc_depth = 3)
```

# Reading and Summarizing Data from the National Hockey League's (NHL) API

## Introduction
This vignette is an introduction on how to read and summarize data using the `tidyverse`. We will be using the National Hockey League's (NHL) API. 

## Load Packages
Before reading and summarizing the data, some necessary packages must be loaded.
```{r packages, echo = TRUE}
library(httr)
library(jsonlite)
library(rmarkdown)
library(knitr)
library(tidyverse)
library(dplyr)
library(readr)
library(DT)
library(lubridate)
library(ggplot2)
library(stringr)
```

## NHL Record API
```{r API reader}
API_info <- function(call){
  base <- "https://records.nhl.com/site/api"
  get_info <- GET(paste0(base, "/", call))
  get_info_text <- content(get_info, as = "text")
  get_info_json <- fromJSON(get_info_text, flatten = TRUE)
  get_info_df <- as.data.frame(get_info_json)
  get_info_tbl <- tbl_df(get_info_df)
  return(get_info_tbl)
}
```

```{r franchise identifers}
franchise_identifer <- function(){
  franchise_IDs <- franchise_info()
  franchise_IDs <- franchise_IDs %>% select(data.fullName, data.id) %>% arrange(data.id)
  colnames(franchise_IDs) <- c("Team Name" , "Franchise ID")
  return(franchise_IDs)
}
```

```{r franchise to ID, error=TRUE, message = TRUE}
franchise_ID <- function(franchise){
  if(is.numeric(franchise)){
    if(franchise > 0 & franchise < 39)
      return(franchise)
    else{
      stop("Please enter a valid franchise ID from 1-38")
    }
  }else{
    ID_df <- API_info("franchise-team-totals")
    if(franchise %in% ID_df$data.teamName){
      ID_num <- ID_df %>% select(data.teamName, data.franchiseId) %>% filter(data.teamName == franchise) %>% select(data.franchiseId) %>% filter(row_number() == 1)
      return(ID_num[[1]][1])
    }else{
      stop("Please enter a valid franchise name")
    }
  }
}
```

```{r team identifers}
team_identifer <- function(){
  team_IDs <- franchise_team_totals()
  team_IDs <- team_IDs %>% select(data.teamName, data.teamId) %>% arrange(data.teamId)
  team_IDs <- team_IDs[seq(1, length(team_IDs$data.teamName), by = 2),]
  colnames(team_IDs) <- c("Team Name" , "Team ID")
  return(team_IDs)
}
```

```{r Team ID, error = TRUE}
team_ID <- function(franchise){
  if(is.numeric(franchise)){
    if(franchise > 0 & franchise <= 58){
      if(franchise %in% c(11,27,31,32,33,34,35,36,38,40,42,44,46,47,48,50,55)){
        stop("No team ID for 11, 27, 31, 32, 33, 34, 35, 36, 38, 40, 42, 44, 46, 47, 48, 50, 55")
      }else{
        return(franchise)
      }
    }else{
      stop("Please enter a valid team ID from 1-55")
    }
  }else{
    ID_df <- API_info("franchise")
    if(franchise %in% ID_df$data.fullName){
      ID_num <- ID_df %>% select(data.fullName, data.mostRecentTeamId) %>% filter(data.fullName == franchise) %>% select(data.mostRecentTeamId) %>% filter(row_number() == 1)
      return(ID_num[[1]][1])
    }else{
      stop("Please enter a valid franchise name")
    }
  }
}
```

### Franchise
```{r NHL Record API franchise}
franchise_info <- function(){
  franchise_df <- API_info("franchise")
  franchise_df <- franchise_df %>% select(data.id, data.mostRecentTeamId, data.fullName, data.teamPlaceName,
                                                  data.teamCommonName, data.teamAbbrev, data.firstSeasonId,
                                                  data.lastSeasonId, total) %>% arrange(data.mostRecentTeamId)
  return(franchise_df)
}
```

### Franchise Team Totals
```{r NHL Record API franchise team totals}
franchise_team_totals <- function(){
  team_df <- API_info("franchise-team-totals")
  team_df$data.activeFranchise <- as_factor(team_df$data.activeFranchise)
  levels(team_df$data.activeFranchise) <- list("No" = 0, "Yes" = 1)
  team_df$data.gameTypeId <- as_factor(team_df$data.gameTypeId)
  levels(team_df$data.gameTypeId) <- list("Regular Season" = 2, "Playoffs" = 3)
  team_df <- team_df %>% select(data.id, data.franchiseId, data.teamName, data.triCode, data.teamId, data.activeFranchise, data.firstSeasonId, data.lastSeasonId, everything()) %>% arrange(data.franchiseId)
  return(team_df)
}
```

### Season Records
```{r NHL Franchise Season Records, error = TRUE}
season_record <- function(franchise = NULL){
  if(is.null(franchise)){
    season_df <- API_info("franchise-season-records")
    season_df <- season_df %>% select(data.id, data.franchiseName, data.franchiseId, everything())
    return(season_df)
  }else{
    season_df <- API_info(paste0("franchise-season-records?cayenneExp=franchiseId=", franchise_ID(franchise)))
    season_df <- season_df %>% select(data.id, data.franchiseName, data.franchiseId, everything())
    return(season_df)
  }
}
```

### Goalie Records
```{r Goalie Records, error = TRUE}
goalie_record <- function(franchise = NULL){
  if(is.null(franchise)){
    goalie_df <- API_info("franchise-goalie-records")
    return(goalie_df)
  }else{
    goalie_df <- API_info(paste0("franchise-goalie-records?cayenneExp=franchiseId=", franchise_ID(franchise)))
    return(goalie_df)
  }
}
```

### Skater Records
```{r Skater Records, error = TRUE}
skater_record <- function(franchise = NULL){
  if(is.null(franchise)){
    skater_df <- API_info("franchise-skater-records")
    return(skater_df)
  }else{
    skater_df <- API_info(paste0("franchise-skater-records?cayenneExp=franchiseId=", franchise_ID(franchise)))
    return(skater_df)
  }
}
```

### Admin History and Retired Numbers
```{r Admin History and Retired Numbers, error = TRUE}
admin_history <- function(franchise = NULL){
  if(is.null(franchise)){
    admin_history_df <- API_info("franchise-detail")
    admin_history_df <- admin_history_df %>% select(data.mostRecentTeamId, data.teamFullName, data.teamAbbrev, data.id, everything())
    return(admin_history_df)
  }else{
    admin_history_df <- API_info(paste0("franchise-detail?cayenneExp=mostRecentTeamId=", team_ID(franchise)))
    admin_history_df <- admin_history_df %>% select(data.mostRecentTeamId, data.teamFullName, data.teamAbbrev, data.id, everything())
    return(admin_history_df)
  }
}
```

## NHL Stats API
```{r NHL Stats API}
team_stats <- function(franchise = NULL){
  base <- "https://statsapi.web.nhl.com/api/v1/teams"
  if(is.null(franchise)){
    url <- paste0(base, "?expand=team.stats")
  }else{
    ID <- team_ID(franchise)
    url <- paste0(base, "/", ID, "?expand=team.stats")
  }
  get_stats <- GET(url)
  get_stats_text <- content(get_stats, as = "text")
  get_stats_json <- fromJSON(get_stats_text, flatten = TRUE)
  get_stats_df <- as.data.frame(get_stats_json)
  get_stats_tbl <- tbl_df(get_stats_df)
  return(get_stats_tbl)
}
```

```{r Stats One Stop}
stats_one <- function(endpoint, franchise = NULL, modifier = NULL){
  base <- "https://statsapi.web.nhl.com/api/v1/"
  if(is.null(endpoint)){
    stop("Please enter valid endpoint")
  }else{
    if(is.null(franchise) & is.null(modifier)){
      url <- paste0(base, endpoint)
      return(url)
    }else if(!is.null(franchise) & is.null(modifier)){
      ID <- team_ID(franchise)
      url <- paste0(base, endpoint, "/", ID)
      return(url)
    }else if(is.null(franchise) & !is.null(modifier)){
      url <- paste0(base, endpoint, modifier)
      return(url)
    }else{
      ID <- team_ID(franchise)
      url <- paste0(base, endpoint, "/", ID, modifier)
      return(url)
    }
  }
}
```

```{r Wrapper function}
NHL_info <- function(endpoint, ...){
  if(endpoint == "franchise"){
    franchise_info()
  }else if(endpoint == "team totals"){
    franchise_team_totals()
  }else if(endpoint == "season records"){
    season_record(...)
  }else if(endpoint == "goalie records"){
    goalie_record(...)
  }else if(endpoint == "skater records"){
    skater_record(...)
  }else if(endpoint == "franchise history"){
    admin_history(...)
  }else if(endpoint == "stats"){
    team_stats(...)
  }else{
    stop("Please enter valid endpoint")
  }
}
```

## NHL Exploratory Data Analysis
### Newer Teams
This contigency table takes a look at the teams from the franchise endpoint of the NHL Records API based on when they joined the NHL. I was curious to see how many teams joined the NHL after I was born, so I categorized all of the teams by older and younger than me. For reference I was born in 1995. Below it can be seen that 6/39 teams joined the league after I was born and are therefore younger than me.

```{r New Teams Table}
franchise_df <- NHL_info("franchise")
franchise_df$data.firstSeasonId <- paste0(substr(franchise_df$data.firstSeasonId,1,4),"/",substr(franchise_df$data.firstSeasonId,5,8))
franchise_df$year.split <- str_split(franchise_df$data.firstSeasonId, "/")
for(j in 1:nrow(franchise_df)){
  if(as.numeric(franchise_df$year.split[[j]][1]) >= 1995){
    franchise_df$new.team[j] <- "Younger than me"
  }else{
    franchise_df$new.team[j] <- "Older than me"
  }
}
franchise_table <- table(franchise_df$new.team)
kable(franchise_table, col.names = c("Team Age", "Count"))
```

### Active Contigency Table
Below is a contigency table that show the number of teams that have made the playoffs after the regular season parsed by the active and inactive franchises. This shows that 5/13 inactive teams and 43/44 active teams ever made the playoffs. By taking the sum of the Regular Season column you can calculate the total number of 57 franchises represented in this endpoint. Based on my knowledge of the NHL, I know that there are only 31 active teams currently in the league so this information is misleading. To take a better look active the active teams that have made the playoffs, see below.

```{r Active Table}
team_totals_df <- NHL_info("team totals")
active_df <- team_totals_df %>% filter(data.gameTypeId == "Regular")
active_table <- table(team_totals_df$data.activeFranchise, team_totals_df$data.gameTypeId)
kable(active_table)
```

### Record Contingency Tables
This is a simple contigency table that shows the number of active teams that have ever made the playoffs after the regular season. There are 31 active teams as seen by the Regular Season row because all teams participate in the regular season and of those 31 teams all of them have made the playoffs before. This data was found by joining information from the NHL Record API and the NHL Stats API.

```{r Playoff Table}
team_totals_df <- NHL_info("team totals")
colnames(team_totals_df)[3] <- "teams.name"
stats_df <- NHL_info("stats")
playoff_df <- left_join(stats_df, team_totals_df)
playoff_table <- table(playoff_df$data.gameTypeId)
kable(playoff_table, col.names = c("Game Type", "Number of Teams"))
```

### Goalie Single Season Records
Below is a summary table of the median values for goalie single season records parsed by active and inactive players. Included is the number of goalies for each player status to show the amount of data being pulled. The data shows that goalies nowadays are a bit more active compared to previous goalies based on the median most shots, but not by a lot. I also wanted to include the median number of saves and median most goals allowed to see if the skill of the goalies have increased or decreased. Based on these values it appears that the skill level of the goalies have remained about the same with a possible slight increase in effectiveness.
```{r WSH Gaolie Summary}
goalie_df <- NHL_info("goalie records")
goalie_df$data.activePlayer <- as_factor(goalie_df$data.activePlayer)
levels(goalie_df$data.activePlayer) <- list("Active" = TRUE, "Inactive" = FALSE)
goalie_game_df <- goalie_df %>% group_by(data.activePlayer) %>% summarise(`Number of Goalies` = n(), `Most Shots` = round(median(data.mostShotsAgainstOneGame, na.rm = TRUE)), `Most Saves` = round(median(data.mostSavesOneGame, na.rm = TRUE)), `Most Goals` = round(median(data.mostGoalsAgainstOneGame, na.rm = TRUE)))
colnames(goalie_game_df)[1] <- "Players Status"
kable(goalie_game_df)
```

### Washington Capitals Penalty Minute Summaries
Below is a summary table of the penalty minutes for the Washington Capitals parsed by position. Included is the number of skaters at each position to show the amount of data being pulled to calculate the averages. Based on the averages the left wings and defensemen spend the most time in the penalty box whereas the Centers seem to play a bit cleaner. However, I found it interesting that the position with the max amount of penalty minutes by far came from a center.
```{r WSH Skater Summary}
WSH_skater_df <- NHL_info("skater records", franchise = "Washington Capitals")
for(k in 1:nrow(WSH_skater_df)){
  WSH_skater_df$full.name[k] <- paste(WSH_skater_df$data.firstName[k], WSH_skater_df$data.lastName[k], sep = " ")
}
WSH_skater_df$data.positionCode <- as_factor(WSH_skater_df$data.positionCode)
levels(WSH_skater_df$data.positionCode) <- list("Left Wing" = "L", "Right Wing" = "R", "Center" = "C", "Defensemen" = "D")
WSH_pen_mins <- WSH_skater_df %>% group_by(data.positionCode) %>% summarise(`Number of Skaters` = n(), `Average Penalty Minutes` = round(mean(data.penaltyMinutes)), `Max Penalty Minutes` = max(data.penaltyMinutes))
colnames(WSH_pen_mins)[1] <- "Position"
kable(WSH_pen_mins)
```

### Barplot of Goalies
Below is a barplot that shows the number of goalies with at least 10 shutouts in a season separated by team in order from most to least. The purpose of this plot is to show the teams that have had the best goalies throughout history. It should be noted that all teams are not represented here meaning that not all teams have had a goalie with 10 or more shutouts in a season. The teams with the most goalies are the Detroit Red Wings and the Boston Bruins with 4 goalies followed by the Toronto Maple Leafs, New York Rangers, and Montréal Canadiens with 3 goalies each.

```{r Goalie Barplot}
library(forcats)
goalie_df <- NHL_info("goalie records")
goalie_df <- goalie_df %>% filter(data.mostShutoutsOneSeason >= 10) %>% arrange(data.mostShutoutsOneSeason)
ggplot(data = goalie_df, aes(y = fct_rev(fct_infreq(data.franchiseName)))) + geom_bar(fill = "blue") + labs(title = "Number of Goalies with 10 or more Shutouts in a Season", x = "Number of Shutouts", y = "Team Name")
```

### Histogram of Win Percentage
```{r percentage histogram}
team_totals_df <- NHL_info("team totals")
team_totals_df <- team_totals_df %>% group_by(data.teamName)
team_totals_df$win.percentage <- team_totals_df$data.wins/(team_totals_df$data.wins + team_totals_df$data.losses)
#ggplot(data = team_totals_df, aes(x = data.teamName, ))
```

### Boxplot of Divsions
Below shows boxplots for season points by division with the individual team points overlaid. Points for each team are calcualted as follows:

  * 2 points for each win
  * 1 point for an overtime/shootout loss
  * 0 points for each loss

The more points a team has the better they are considered to be. From the visual below it can be seen that the MassMutaul East was the best division on average, however, the best teams from the Discover Central and Honda West were better than their best teams. 

```{r Division Boxplot}
team_stats_df_1 <- team_stats()
team_stats_df <- team_stats_df_1 %>% select(teams.name, teams.division.name, teams.teamStats)
points <- c()
for(i in 1:nrow(team_stats_df)){
  points <- c(points, ((team_stats_df[[3]][[i]])[[1]][[1]])$stat.pts[1])
}
team_stats_df <- team_stats_df[-32,]
team_stats_df <- cbind(team_stats_df, points)
team_stats_df <- team_stats_df %>% select(-teams.teamStats)
team_stats_df$teams.division.name <- as_factor(team_stats_df$teams.division.name)
team_stats_df$points <- as.numeric(as.character(team_stats_df$points))
ggplot(data = team_stats_df, aes(x = teams.division.name, y = points)) + geom_boxplot() + geom_point(aes(color = teams.division.name), position = "jitter") + labs(title = "Boxplots of Season Points by Division", x = "Division", y = "Season Points") + scale_color_discrete(name = "Divisions")
```

